Application of Generalized Additive Models for long-term and seasonal trend analysis of water quality: South San Francisco Bay case study

### 2020-04-20 manu_draft.Rmd
cloern reminder for intro
getting through hypothesis tests for gams

@@

All models were compared using standard summary statistics describing overall model fit to the observed data.  In addition to the GCV score, models were comapred using the Akaike Information Criterion (AIC) and $R^2$ summmary statistics.  Comparisons were made between the different GAMs in equations \@ref(eq:gamzero) to \@ref(eq:gamsix) and the transformation used to define the $g()$ function for chlorophyll at each station. Model predictions were based on the standard model output from the *predict()* function from the mgcv package.  A long-term trend independent of seasonal variation was also estimated for each model by subtracting the seasonal terms (i.e., $s_1(doy)$ and $ti_1(doy, year)$) from the predictions. Back-transformation of results were obtained by exponentiation if the log-transformed variable was used or using the *InvBoxCox()* function from the forecast package if the Box-Cox transformation was used.

## Long-term and seasonal trend analysis

>>>>

Cloern et al. 2020 [link](https://link.springer.com/article/10.1007/s12237-020-00737-w)

@@

All models were compared using standard summary statistics describing overall model fit to the observed data.  In addition to the GCV score, models were compared using the Akaike Information Criterion (AIC) and $R^2$ summmary statistics.  Comparisons were made between the different GAMs in equations \@ref(eq:gamzero) to \@ref(eq:gamsix) and for the transformations used to define the $g()$ function for chlorophyll at each station. Model predictions were based on the standard model output from the *predict()* function from the mgcv package.  A long-term trend independent of seasonal variation was also estimated for each model by subtracting the seasonal terms (i.e., $s_1(doy)$ and $ti_1(doy, year)$) from the predictions. Back-transformation of results were obtained by exponentiation if the log-transformed variable was used or using the *InvBoxCox()* function from the forecast package if the Box-Cox transformation was used. For hypothesis tests that required estimates of central tendency and confidence intervals for mean values, back-transformations were based on an exponentiation or inverse function of the transformed response variable as before, with the addition of a correction factor.  This bias-correction factor was simply the variance of the model residuals divided by two for variables in log-space and correction with a forecast variance estimate as specified in the *InvBoxCox()* function from the forecast package.

## Trend tests on GAM results

Each of the GAM structures described in equations \@ref(eq:gamzero) to \@ref(eq:gamsix) provide different approaches to quantify long-term variation in annual and seasonal changes.  Although useful as a general approach to view functional forms for different time series, additional information regarding whether an observed change is statistically significant can be useful to inform environmental management decisions.  We present two methods for performing trend tests on model results that address different changes that may be of management interest.  First, a method for evaluating the percent change in annual averages between two time periods is briefly summarized.  This approach is similar to that presented in Murphy et al. (2019).  Second, a method for evaluating means and confidence intervals from GAM results for seasonal periods is presented, including an estimate of direction and rate of seasonal change over time.

The ability to compare time periods of interest has importance for evaluating whether a significant change has occurred in relation to perceived trends in observed data.  For example, the noticeable change in SFE water quality in South Bay starting from the 1990s and persisting through 2005 was of considerable interest by environmental managers given the apparent departure from long-term conditions to a state that suggested a trend towards more eutrophic conditions.  However, statistical methods to evaluate this change that were able to quantify the temporal variation and to provide a statistically robust estimate of the change were not available.  The use of GAMs in parallel with hypothesis testing can be used for this purpose. Methods proposed in Murphy et al. (2019) and similarly applied in Yang and Moyer (2020) were used and are briefly summarized herein.

For two time periods of interest, a trend can be estimated as the difference in means from the GAM results between the periods, the percent change, and an appropriate p-value descibing the probability that the change is due to change alone.  Murphy et al. (2019) defined the two time periods as "base" and "test" periods, where the former represents a baseline condition earlier in the time series as a basis of comparison to the more recent "test" period.  It was further suggested the the base and test periods include a minimum of two years if evaluating years at the beginning and end of a time series to account for instability of GAM estimates at the edges of the independent variable space.  The differences are estimated as:

\begin{equation}
\text{Difference} = Z_d\hat{\beta}
(\#eq:diffest)
\end{equation}

where $Z_d$ is a point estimate of the differences of the predictions between the base and test years and $\hat{\beta}$ is the estimated parameter vector from the fitted GAM.  The standard error of this difference is then:

\begin{equation}
\text{se}_{\text{diff}} = \sqrt{Z_d\hat{\Sigma}_\beta Z_d^T}
(\#eq:sediffeest)
\end{equation}

defined as the square root of the product of $Z_d$, the estimated variance-covariance matrix of parameter vector for the fitted GAM $\hat{\Sigma}_\beta$, and the transposition of $Z_d$.  The trend is considered significant if the appropriate confidence interval at a specified $\alpha$ does not include zero.

The second hypothesis test evaluates changes in seasonal estimates between years using results from a fitted GAM.  This test provides both an estimate of a seasonal mean and its associated confidence interval and if the rate of change in the seasonal means has varied over a specific period interest.  For example,
